---
title: "Statistical analysis of significantly mutated genes"
author: "Collin Tokheim"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Statistical analysis of significantly mutated genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Identifying genes with more mutations then expected has been central methodology for identifying putative cancer driver genes in exome sequencing studies of cancer samples. Identifying significantly mutated genes fundamentally relies on estimating a background mutation rate. Mutation rate varies over more than 2 orders of magnitude providing a substantial statistical estimation challenge. Analysis not accounting for the uncertainty in mutation rate yields overly optimistic assessments. In this package, we examine statistical power (either with known or uncertain mutation rate) and false positives induced by unaccounted variation in mutation rate.

## Sample Size Calculation

An important aspect of designing cancer exome seqeuncing studies is to determine how many cancer samples are required for sufficient power to detect significantly mutated genes present at a certain prevalence.

### Assuming a known mutation rate

In general the mutation rate is not precisely known, but could be assumed to be known for the sake of power calculations. This results in an overly optimistic assessment of the required number of cancer samples. In the known mutation rate scenario, an exact binomial power calculation is performed.

```{r, fig.show='hold'}
library(cancerSeqStudy)

# setup parameters
samp.sizes <- seq(100, 4000, by=100)
desired.power <- .9
eff.size <- .02  # fraction of samples above background
mut.rate <- 1e-5
alpha.level <- 5e-6  # roughly a bonferoni corrected significance level

binomRequiredSampleSize(desired.power, mut.rate, samp.sizes, eff.size, alpha.level)
```

### Accounting for uncertain mutation rate

Adjusting for uncertainty in mutation rate better represents the actual required number of sequenced samples. To handle uncertain mutation rates (with a certain coefficient of variation), a beta-binomial power test is performed.


```{r, fig.show='hold'}
# setup parameters
cv <- .2  # coefficient of variation for mutation rate

bbdRequiredSampleSize(desired.power, mut.rate, cv, samp.sizes, eff.size)
```

Notice the minimum required samples raised from 1,500 to 3,500 by accounting for uncertainty of mutation rate with a coefficient of variation of .2.

## Calculating powered effect size

If you already have a certain number of samples, often it is helpful to understand the extent to which rare significantly mutated genes are characterized. 


```{r, fig.show='hold'}
# setup parameters
possible.eff.sizes <- seq(.01, .2, by=.01)  # fraction of samples above background
num.samples <- 1000  # number of samples in study

bbdPoweredEffectSize(possible.eff.sizes, desired.power, mut.rate, cv, num.samples, alpha.level)
```

In this example, drivers present in 4% of samples above background mutation rate have sufficient power to be detected.

## Expected false positives



## Systematically examining power and false postives

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
